---
title: "CM580A3 - Intro qCMB - R Databases (Part 1)"
author: "Jessica Hill"
date: '`r Sys.Date()`'
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = FALSE)
```

***

Lets start off by making sure all of the example data sets are downloaded.

Once the data is downloaded and or the repository is cloned to your desktop using GitHub desktop, lets find our current working directory and then set it as our working directory. 

This will make it easier for us to point to files during the lesson. 

```{r}
getwd() # find our current working directory 

#setwd("/Users/jessicahill/Library/CloudStorage/OneDrive-Colostate/JESSICA/CSU/BMB/ONISH_LAB/PROJECTS/PROGRAMMING/Intro_qCMB_class/2023_qCMB_class") # set the directory we want as our current working directory 
```

*** 

### Quick Review

> Basic data types in R:

* Logical (aka Boolean)
    + Can only have two values. 
    + For example: TRUE/FALSE, 0/1 (integer, but can be coerced to Boolean), Yes/No (concept only, can’t be coerced to values)
    
```{r}
class(TRUE)
```
```{r}
# Useful functions

#class() - what kind of object is it (high-level)?
#typeof() - what is the object’s data type (low-level)?
#length() - how long is it? What about two dimensional objects?
#attributes() - does it have any metadata?
```

* Numeric 
    + All real numbers.
    + For example: 3, 3.4
    
```{r}
class(3)

class(3.4)
typeof(3.4) # is actually a double precision floating point value
```

* Integers  
    + Integers are a special kind of numeric data.
    + For example: 1, 2, 3 
    
```{r}
class(3L) 
# the L signifies that the number is an integer 
# we can compare the class() and typeof() for both 3 and 3L

y <- 1:10 # this notation (1:10) is designating a range of numbers, 1 through 10, and assigning them to the variable y.
class(y)
```   
  
* Complex
    + Specify imaginary values in R. 
    + The suffix i is used to specify the imaginary part. 
    + For example:
    
```{r}
class(1+4i) 
class(2i)
```         
  
* Character
    + Specify character or string values in a variable. 
    + A string is a set of characters. 
    + You can use '' or "" to represent strings. "A" is a single character and "Apple" is a string
    + For example: 
    
```{r}
class("some text")
class("1")
class('a')
```            

* Raw
    + Specifies values as raw bytes. 
    + The bytes are represented in hexadecimal. 
    + For example: 
    
```{r}
raw_variable <- charToRaw("Welcome to Programiz in English")

class(raw_variable)
print(raw_variable)

# why might this be useful? 

```

[what is hexadecimal?](https://www.techtarget.com/whatis/definition/hexadecimal)

---

> Basic data structures in R:

* Dataframe
    + 2-D collection of elements of different data types. Is truly a special type of list where every element of the list has the same length. 
    + Each column consists of the value of 1 variable.
    + Each row consists of a value set from each column. 
    + For example:
    
```{r}
data_df <- data.frame(id = letters[1:10], x = 1:10, y = 11:20)

print(data_df)
``` 
```{r}
# Useful data frame functions

#head() - shows first 6 rows
#tail() - shows last 6 rows
#dim() - returns the dimensions of data frame (i.e. number of rows and number of columns)
#nrow() - number of rows
#ncol() - number of columns
#str() - structure of data frame - name, type and preview of data in each column
#names() or colnames() - both show the names attribute for a data frame
#sapply(dataframe, class) - shows the class of each column in the data frame
```

* Vectors 
    + 1-D list of the SAME DATA type.
    + There are 6 types of atomic vectors: logical, integer, character, raw, double, and complex.
    + For example: 
    
```{r}
organism <- c('human', 'mouse', 'worm', 'yest', 'maize')

class(organism)
print(organism)


numbers <- c(1, 2, 3, 1, 2, 4)

class(numbers)
print(numbers)


logical <- c(TRUE, FALSE, FALSE, TRUE)

class(logical)
print(logical)
```
* Lists
    + 1-D or 2-D list containing DIFFERENT types of data.
    + Necessary for non-tabular data.
    + For example: 
    
```{r}         
random <- list("human", 1, 3, TRUE, 'worm', c(1,2,3), c(TRUE, FALSE, TRUE))

class(random)
print(random)
```     

* Matrices
    + 2-D collection of elements of the SAME DATA type. 
    + Usually contains numeric values in order to perform mathematical operations (i.e., least squares). 
    + For example:
    
```{r}
mx <- matrix(
c(1,2,3,4,5,6,7,8),
nrow = 4,
ncol = 2,
byrow = TRUE)     

class(mx)
print(mx)
``` 

* Array
    + Is a vector with 1 or more dimensions.
        + A 1-D array is really just a vector
        + A 2-D array is really just a matrix
    + Multi-dimensional data is stored in the form of an n-dimensional matrix.
    + For example: 
    
```{r}
v1 <- c(1,2,3)
v2 <- c(4,5,6,7,8,9)
result <- array(c(v1,v2),dim = c(3,3,2))

class(result)
print(result)
```     

```{r}
v1 <- c(1,2,3)
v2 <- c(4,5,6,7,8,9)
col.names <- c("Item","Serial","Size")
row.names <- c("Server","Network","Firewall")
matrix.names <- c("DataCenter EU","DataCenter US")
result2 <- array(c(v1,v2),dim = c(3,3,2),dimnames = list(row.names,col.names,matrix.names))

print(result2)
```
    
* Factor 
    + Are data objects that are used in order to categorize and store data as levels.
        + A base R way to “group_by” (as in tidyverse) value levels.
    + They can be strings or integers. 
    + For example:
    
```{r}
gender <- factor(c("female", "female", "male", "female", "male")) 

class(gender)
levels(gender) # you can check how many or what levels a factor variable has 
```   

***

### Importing/Exporting data in R

> Considerations

Before importing data into R, we need to check how our data is formatted.

* Does your data file have row and column names? 
    + If not, consider adding them. 
    + If so, are they using good or bad naming conventions? 
        
```
# Helpful tips

- Avoid names with blank spaces. 
    Good column names: Long_jump or Long.jump
    Bad column name: Long jump

- Avoid names with special symbols: ?, $, *, +, #, (, ), -, /, }, {, |, >, <
    Only underscore can be used.

- Avoid beginning variable names with a number. Use letter instead. 
    Good column names: sport_100m or x100m 
    Bad column name: 100m

- Each variable has its own column. Each row is an observation.
    All data in a column has the same type (e.g. it is all numerical data, or it is all text data)

- Column names must be unique and describe the variable contained within that column. Duplicated names are not allowed.

- No blank columns or blank rows between data

- R is case sensitive. This means that Name is different from Name or NAME.

- Delete any comments in your file

- Replace missing values by NA (for not available)

- If you have a column containing date, use the four digit format. 
    Good format: 01/01/2016 
    Bad format: 01/01/16
    
- Data are consistent (e.g. if a binary variable can take values ‘Yes’ or ‘No’ then only these two values are allowed, with no alternatives such as ‘Y’ and ‘N’)
```

* Check out this resource on recognizing and resolving common spreadsheet formatting problems. 
    + [Formatting Problems](https://datacarpentry.org/spreadsheet-ecology-lesson/02-common-mistakes/)

```
# Key take aways from the above resource

- Avoid using multiple tables within one spreadsheet.
- Avoid spreading data across multiple tabs.
- Record zeros as zeros. Dont leave them blank.
- Use an appropriate null value to record missing data.
- Don’t use formatting to convey information or to make your spreadsheet look pretty.
- Place comments in a separate column.
- Record units in column headers.
- Include only one piece of information in a cell.
- Avoid spaces, numbers and special characters in column headers.
- Avoid special characters in your data.
- Record metadata in a separate plain text file.
```

We also need to make sure that the data is saved in a file format that R can recognize. 

* What type of file formats can R recognize?
    + Quite a lot actually. 
    + Here are some resources to check out below 
        + [Import, Export, and Convert Data Files](https://cran.r-project.org/web/packages/rio/vignettes/rio.html)
        + [R Data Import/Export](https://cran.r-project.org/doc/manuals/r-release/R-data.html) 

* However, the file formats you will most likely work with are .csv, .tsv or .txt, .xlsx, JSON, Database, XML/HTML 

---

> How can we import data into R? 

* Many ways to import data into R, but usually we use a function that is somewhat specific to the file type we are trying to import. 

* Lets check out some examples below: 

Read in a .csv file, using 'read.csv'
```{r}
data1 <- read.csv('./Example_Data/heart_data.csv')
head(data1, 5)
```

Now read in the same .csv file, but using 'read.table'
```{r}
data2 <- read.table('./Example_Data/heart_data.csv', sep=",", header = 1)
head(data2, 5)
```

Read in a large zipped .csv file 
```
file <- unz("./Example_Data/US_Accidents.csv.zip", "US_Accidents.csv")
data3 <- read.table(file, header=TRUE, sep=",",nrow=10000)
data3

# need to cut this code chunk out to render the document, otherwise it wont
```

Read in a .txt file, using 'read.delim' 
```{r}
data4 <- read.delim("./Example_Data/TulsaJun20_2020.txt",header = F)
head(data4, 10)
```

Read in an Excel file
```{r}
#install.packages("tidyverse") 
library(readxl) 

Data <- read_excel("./Example_Data/HNP_Stats.xlsx", sheet = 1)
head(Data, 5)

Country <- read_excel("./Example_Data/HNP_Stats.xlsx", sheet = 2)
head(Country, 5)

Series <- read_excel("./Example_Data/HNP_Stats.xlsx", sheet = 3)
head(Series, 5)
```

Read in a JSON (JavaScript Object Notation) file
```{r}
#install.packages("rjson")
library(rjson)

JsonData <- fromJSON(file = './Example_Data/iris.json')
print(JsonData[1:10])

# to convert the JSON data into an R dataframe, use 'as.data.frame()' function in the data.table package
#install.packages("data.table") 
library(data.table)

data5 = as.data.frame(JsonData[1:5])
data5
```

Import data from a Database using SQL
```{r}
#install.packages("DBI")
library(DBI)
library(RSQLite)

# load the database using dbConnect()
conn <- dbConnect(RSQLite::SQLite(), "./Example_Data/basketball.sqlite")

# display the names of the tables
dbListTables(conn)

# run a query and display the results, using the dbGetQuery() function. Just add a SQLite connection object and SQL query as a string.  
print(dbGetQuery(conn, "SELECT * FROM player"))

data6 = dbGetQuery(conn, "SELECT * FROM inactive_players")
data6
```

Import data from XML files
[XML Repository with example data](https://www.w3schools.com/xml/)
```{r}
#install.packages("XML", "xml2")
library(XML)
library(xml2)

breakfast_menue <- read_xml("https://www.w3schools.com/xml/simple.xml")

breakfast_menue_parse <- xmlParse(breakfast_menue)
breakfast_menue_parse

breakfast_menue_nodes= getNodeSet(breakfast_menue_parse, "//food")
breakfast_menue_nodes

data7 <- xmlToDataFrame(breakfast_menue_parse, nodes=breakfast_menue_nodes)
head(data7)
```

Import data from HTML files
```{r}
#install.packages("RCurl", "XML")
library(RCurl)
library(XML)

url <- getURL("https://en.wikipedia.org/wiki/Brazil_national_football_team")
tables <- readHTMLTable(url) # extract all HTML tables from the website
data8 <- tables[23]
data8$`NULL`
print(data8)

# OR

#install.packages("rvest")
library(rvest)

url <- "https://en.wikipedia.org/wiki/Brazil_national_football_team"
file<-read_html(url)
tables<-html_nodes(file, "table")
data9 <- html_table(tables[25])
View(data9)
```

Import binary data (i.e., an image)
```{r}
#install.packages("tiff")
library(tiff)
brain <- readTIFF("./Example_Data/brain.tif")

## set up the plot region:
op <- par(bg = "thistle")
plot(c(100, 250), c(300, 450), type = "n", xlab = "", ylab = "")
image <- as.raster(brain)
rasterImage(image, 100, 300, 250, 450, angle = 0, interpolate = FALSE)
par(op)

# OR

#install.packages("raster")
#install.packages("rgdal")
#install.packages("sp")
library(raster)
library(rgdal)
library(sp)

multiRasterB <- brick(paste0("./Example_Data/brain.tif"))
plotRGB(multiRasterB,r=1,g=2,b=3, stretch="lin")
```

Also, check out these resources for more information and examples: 

[R data import tutorial](https://www.datacamp.com/tutorial/r-data-import-tutorial)

[Image raster data in R](https://www.neonscience.org/resources/learning-hub/tutorials/image-raster-data-r)

---

> How can we export data from R?

* There are a several ways to export data from R

* Lets check out some examples below: 

But first we need to read in some data
```{r}
#install.packages("tidyverse")
library(tidyverse)

winequality_df <- read.csv("./Example_Data/winequality-red.csv", header = T)

new_winequality_df <- winequality_df %>% filter(quality > 6)
```

Export a dataframe to a .txt file, using write.table 
```{r}
write.table(new_winequality_df, file = './Example_Data/new_winequality_df.txt', col.names = TRUE,
             row.names = FALSE, sep = "\t")
```

Export a dataframe to a .csv file, using write.table
```{r}
write.table(new_winequality_df, file = './Example_Data/new_winequality_df.csv', col.names = TRUE,
             row.names = FALSE, sep = ",")
```

Now export the same dataframe to a .csv file, using write.csv 
```{r}
write.csv(new_winequality_df, file = './Example_Data/new_winequality_df2.csv', row.names = FALSE)
```

An alternative to write.table (is meant to be faster)
```{r}
#install.packages("data.table")
library(data.table)

fwrite(new_winequality_df, file = './Example_Data/new_winequality_df2.txt', sep = "\t")

fwrite(new_winequality_df, file = './Example_Data/new_winequality_df3.csv')
```

Another alternative to write.table
```{r}
library(readr)
write_tsv(new_winequality_df, path = './Example_Data/new_winequality_df3.txt')

write_csv(new_winequality_df, path = './Example_Data/new_winequality_df4.csv')
```

---

> Are all data file formats handled equally?

Lets check the reading speeds of the different functions 
```{r}
# Use system.time() to time how fast R processes an expression. 
# This function is especially useful for comparing the speed of two coding options.

system.time(
  read.csv("./Example_Data/winequality-red.csv", header = T)
)
```

```{r}
system.time({
  read.table("./Example_Data/winequality-red.csv", header = T, sep = ",");
  #read.delim("./Example_Data/new_winequality_df.txt", header = T, sep = "\t")
})
```

What about writing speeds? 

* As a little home practice: 
    + Check to see which functions write data files faster 
    + Also see if particular file types write faster then others

> What if we have a really large data file?

* May run into R memory issues

* Check out the below resource for some extra reading into this topic 
    + [Out of memory strategies](https://bookdown.org/rdpeng/RProgDA/working-with-large-datasets.html)

---

Let's move over to Part 2!

[Part 2](R_Databases_Part2.md)
