---
title: "CM580A3 - Intro qCMB - R Databases (Part 1)"
author: "Jessica Hill"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = FALSE)
```

***

### Quick Review

> Everything in R is an object!

> Basic data types in R:

1. Logical (aka Boolean)
    + Can only have two values. 
    + For example: TRUE/FALSE, 0/1, Yes/No
```{r}
class(TRUE)
```
```{r}
# Useful functions

#class() - what kind of object is it (high-level)?
#typeof() - what is the object’s data type (low-level)?
#length() - how long is it? What about two dimensional objects?
#attributes() - does it have any metadata?
```

2. Numeric 
    + All real numbers.
    + For example: 3, 3.4
```{r}
class(3)

class(3.4)
typeof(3.4) # is actually a double precision floating point value
```

3. Integers  
    + Integers are a special kind of numeric data.
    + For example: 1, 2, 3 
```{r}
class(3L)

y <- 1:10
class(y)
```   
  
4. Complex
    + Specify imaginary values in R. 
    + The suffix i is used to specify the imaginary part. 
    + For example:
```{r}
class(1+4i) 
class(2i)
```         
  
5. Character
    + Specify character or string values in a variable. 
    + A string is a set of characters. 
    + You can use '' or "" to represent strings. "A" is a single character and "Apple" is a string
    + For example: 
```{r}
class("some text")
class("1")
class('a')
```            
   
6. Raw
    + Specifies values as raw bytes. 
    + The bytes are represented in hexadecimal. 
    + For example: 
```{r}
raw_variable <- charToRaw("Welcome to Programiz in English")

class(raw_variable)
print(raw_variable)
```

[what is hexadecimal?](https://www.techtarget.com/whatis/definition/hexadecimal)

> Basic data structures in R:
   
1. Vectors 
    + 1-D list of the same data type.
    + There are 6 types of atomic vectors: logical, integer, character, raw, double, and complex.
    + For example: 
```{r}
organism <- c('human', 'mouse', 'worm', 'yest', 'maize')

class(organism)
print(organism)


numbers <- c(1, 2, 3, 1, 2, 4)

class(numbers)
print(numbers)


logical <- c(TRUE, FALSE, FALSE, TRUE)

class(logical)
print(logical)
```

2. Lists
    + 1-D or 2-D list containing different types of data.
    + For example: 
```{r}         
random <- list("human", 1, 3, TRUE, 'worm', c(1,2,3), c(TRUE, FALSE, TRUE))

class(random)
print(random)
```     

3. Matrices
    + 2-D collection of elements of the same data type. 
    + Usually contains numeric values in order to perform mathematical operations. 
    + For example:
```{r}
mx <- matrix(
c(1,2,3,4,5,6,7,8),
nrow = 4,
ncol = 2,
byrow = TRUE)     

class(mx)
print(mx)
``` 

4. Array
    + Is a vector with 1 or more dimensions.
        + A 1-D array is really just a vector
        + A 2-D array is really just a matrix
    + Multi-dimensional data is stored in the form of an n-dimensional matrix.
    + For example: 
```{r}
v1 <- c(1,2,3)
v2 <- c(4,5,6,7,8,9)
result <- array(c(v1,v2),dim = c(3,3,2))

class(result)
print(result)
```     

```{r}
v1 <- c(1,2,3)
v2 <- c(4,5,6,7,8,9)
col.names <- c("Item","Serial","Size")
row.names <- c("Server","Network","Firewall")
matrix.names <- c("DataCenter EU","DataCenter US")
result2 <- array(c(v1,v2),dim = c(3,3,2),dimnames = list(row.names,col.names,matrix.names))

print(result2)
```
    
5. Dataframe
    + 2-D collection of elements of different data types. Is truly a special type of list where every element of the list has the same length. 
    + Each column consists of the value of 1 variable.
    + Each row consists of a value set from each column. 
    + For example:
```{r}
data_df <- data.frame(id = letters[1:10], x = 1:10, y = 11:20)

print(data_df)
``` 
```{r}
# Useful data frame functions

#head() - shows first 6 rows
#tail() - shows last 6 rows
#dim() - returns the dimensions of data frame (i.e. number of rows and number of columns)
#nrow() - number of rows
#ncol() - number of columns
#str() - structure of data frame - name, type and preview of data in each column
#names() or colnames() - both show the names attribute for a data frame
#sapply(dataframe, class) - shows the class of each column in the data frame
```

6. Factor 
    + Are data objects that are used in order to categorize and store data as levels.
    + They can be strings or integers. 
    + For example:
```{r}
gender <- factor(c("female", "female", "male", "female", "male")) 

class(gender)
levels(gender) # you can check how many or what levels a factor variable has 
```   
    
***

### Importing/Exporting date in R

> Considerations

Before importing data into R, we need to check how our data is formatted.

* Does your data file have row and column names? 
    + If not, consider adding them. 
    + If so, are they using good or bad naming conventions? 
        
```
# Helpful tips

- Avoid names with blank spaces. 
    Good column names: Long_jump or Long.jump
    Bad column name: Long jump

- Avoid names with special symbols: ?, $, *, +, #, (, ), -, /, }, {, |, >, <
    Only underscore can be used.

- Avoid beginning variable names with a number. Use letter instead. 
    Good column names: sport_100m or x100m 
    Bad column name: 100m

- Each variable has its own column. Each row is an observation.
    All data in a column has the same type (e.g. it is all numerical data, or it is all text data)

- Column names must be unique and describe the variable contained within that column. Duplicated names are not allowed.

- No blank columns or blank rows between data

- R is case sensitive. This means that Name is different from Name or NAME.

- Delete any comments in your file

- Replace missing values by NA (for not available)

- If you have a column containing date, use the four digit format. 
    Good format: 01/01/2016 
    Bad format: 01/01/16
    
- Data are consistent (e.g. if a binary variable can take values ‘Yes’ or ‘No’ then only these two values are allowed, with no alternatives such as ‘Y’ and ‘N’)
```

* Check out this resource on recognizing and resolving common spreadsheet formatting problems. 
    + [Formatting Problems](https://datacarpentry.org/spreadsheet-ecology-lesson/02-common-mistakes/)

```
# Key take aways from the above resource

- Avoid using multiple tables within one spreadsheet.
- Avoid spreading data across multiple tabs.
- Record zeros as zeros. Dont leave them blank.
- Use an appropriate null value to record missing data.
- Don’t use formatting to convey information or to make your spreadsheet look pretty.
- Place comments in a separate column.
- Record units in column headers.
- Include only one piece of information in a cell.
- Avoid spaces, numbers and special characters in column headers.
- Avoid special characters in your data.
- Record metadata in a separate plain text file.
```

We also need to make sure that the data is saved in a file format that R can recognize. 

* What type of file formats can R recognize?
    + Quite a lot actually. 
    + Here are some resources to check out below 
        + [Import, Export, and Convert Data Files](https://cran.r-project.org/web/packages/rio/vignettes/rio.html)
        + [R Data Import/Export](https://cran.r-project.org/doc/manuals/r-release/R-data.html) 

* However, the file formats you will most likely work with are .csv, .tsv or .txt, .xlsx, JSON, Database, XML/HTML 

> How can we import data into R? 

* Many ways to import data into R, but usually we use a function that is somewhat specific to the file type we are trying to import. 

* Lets check out some examples below: 

Read in a .csv file, using 'read.csv'
```{r}
data1 <- read.csv('~/Library/CloudStorage/OneDrive-Colostate/JESSICA/CSU/BMB/ONISH_LAB/PROJECTS/PROGRAMMING/Intro_qCMB_class/2023_qCMB_class/Example_Data/heart_data.csv')
head(data1, 5)
```

Now read in the same .csv file, but using 'read.table'
```{r}
data2 <- read.table('~/Library/CloudStorage/OneDrive-Colostate/JESSICA/CSU/BMB/ONISH_LAB/PROJECTS/PROGRAMMING/Intro_qCMB_class/2023_qCMB_class/Example_Data/heart_data.csv', sep=",", header = 1)
head(data2, 5)
```

Read in a large zipped .csv file 
```{r}
file <- unz("~/Library/CloudStorage/OneDrive-Colostate/JESSICA/CSU/BMB/ONISH_LAB/PROJECTS/PROGRAMMING/Intro_qCMB_class/2023_qCMB_class/Example_Data/US_Accidents.csv.zip", "US_Accidents.csv")
data3 <- read.table(file, header=TRUE, sep=",",nrow=10000)
data3
```

Read in a .txt file, using 'read.delim' 
```{r}
data4 <- read.delim("~/Library/CloudStorage/OneDrive-Colostate/JESSICA/CSU/BMB/ONISH_LAB/PROJECTS/PROGRAMMING/Intro_qCMB_class/2023_qCMB_class/Example_Data/TulsaJun20_2020.txt",header = F)
head(data4, 10)
```

Read in an Excel file
```{r}
install.packages("tidyverse") 
library(readxl) 

Data <- read_excel("~/Library/CloudStorage/OneDrive-Colostate/JESSICA/CSU/BMB/ONISH_LAB/PROJECTS/PROGRAMMING/Intro_qCMB_class/2023_qCMB_class/Example_Data/HNP_Stats.xlsx", sheet = 1)
head(Data, 5)

Country <- read_excel("~/Library/CloudStorage/OneDrive-Colostate/JESSICA/CSU/BMB/ONISH_LAB/PROJECTS/PROGRAMMING/Intro_qCMB_class/2023_qCMB_class/Example_Data/HNP_Stats.xlsx", sheet = 2)
head(Country, 5)

Series <- read_excel("~/Library/CloudStorage/OneDrive-Colostate/JESSICA/CSU/BMB/ONISH_LAB/PROJECTS/PROGRAMMING/Intro_qCMB_class/2023_qCMB_class/Example_Data/HNP_Stats.xlsx", sheet = 3)
head(Series, 5)
```

Read in a JSON (JavaScript Object Notation) file
```{r}
install.packages("rjson")
library(rjson)

JsonData <- fromJSON(file = '~/Library/CloudStorage/OneDrive-Colostate/JESSICA/CSU/BMB/ONISH_LAB/PROJECTS/PROGRAMMING/Intro_qCMB_class/2023_qCMB_class/Example_Data/iris.json')
print(JsonData[1:10])

# to convert the JSON data into an R dataframe, use 'as.data.frame()' function in the data.table package
install.packages("data.table") 
library(data.table)

data5 = as.data.frame(JsonData[1:5])
data5
```

Import data from a Database using SQL
```{r}
install.packages("DBI")
library(DBI)
library(RSQLite)

# load the database using dbConnect()
conn <- dbConnect(RSQLite::SQLite(), "~/Library/CloudStorage/OneDrive-Colostate/JESSICA/CSU/BMB/ONISH_LAB/PROJECTS/PROGRAMMING/Intro_qCMB_class/2023_qCMB_class/Example_Data/basketball.sqlite")

# display the names of the tables
dbListTables(conn)

# run a query and display the results, using the dbGetQuery() function. Just add a SQLite connection object and SQL query as a string.  
print(dbGetQuery(conn, "SELECT * FROM player"))

data6 = dbGetQuery(conn, "SELECT * FROM inactive_players")
data6
```

Import data from XML files
[XML Repository with example data](https://www.w3schools.com/xml/)
```{r}
install.packages("XML", "xml2")
library(XML)
library(xml2)

breakfast_menue <- read_xml("https://www.w3schools.com/xml/simple.xml")

breakfast_menue_parse <- xmlParse(breakfast_menue)
breakfast_menue_parse

breakfast_menue_nodes= getNodeSet(breakfast_menue_parse, "//food")
breakfast_menue_nodes

data7 <- xmlToDataFrame(nodes=breakfast_menue_nodes)
head(data7)
```

Import data from HTML files
```{r}
install.packages("RCurl", "XML")
library(RCurl)
library(XML)

url <- getURL("https://en.wikipedia.org/wiki/Brazil_national_football_team")
tables <- readHTMLTable(url) # extract all HTML tables from the website
data8 <- tables[23]
data8$`NULL`
print(data8)

# OR

install.packages("rvest")
library(rvest)

url <- "https://en.wikipedia.org/wiki/Brazil_national_football_team"
file<-read_html(url)
tables<-html_nodes(file, "table")
data9 <- html_table(tables[25])
View(data9)
```

Import binary data (i.e., an image)
```{r}
install.packages("tiff")
library(tiff)
brain <- readTIFF("~/Library/CloudStorage/OneDrive-Colostate/JESSICA/CSU/BMB/ONISH_LAB/PROJECTS/PROGRAMMING/Intro_qCMB_class/2023_qCMB_class/Example_Data/brain.tif")

## set up the plot region:
op <- par(bg = "thistle")
plot(c(100, 250), c(300, 450), type = "n", xlab = "", ylab = "")
image <- as.raster(brain)
rasterImage(image, 100, 300, 250, 450, angle = 0, interpolate = FALSE)
par(op)

# OR

install.packages("raster")
install.packages("rgdal")
install.packages("sp")
library(raster)
library(rgdal)
library(sp)

multiRasterB <- brick(paste0("~/Library/CloudStorage/OneDrive-Colostate/JESSICA/CSU/BMB/ONISH_LAB/PROJECTS/PROGRAMMING/Intro_qCMB_class/2023_qCMB_class/Example_Data/brain.tif"))
plotRGB(multiRasterB,r=1,g=2,b=3, stretch="lin")
```

Also, check out these resources for more information and examples: 

[R data import tutorial](https://www.datacamp.com/tutorial/r-data-import-tutorial)

[Image raster data in R](https://www.neonscience.org/resources/learning-hub/tutorials/image-raster-data-r)

> How can we export data from R?

* There are a several ways to export data from R

* Lets check out some examples below: 

But first we need to read in some data
```{r}
install.packages("tidyverse")
library(tidyverse)

winequality_df <- read.csv("~/Library/CloudStorage/OneDrive-Colostate/JESSICA/CSU/BMB/ONISH_LAB/PROJECTS/PROGRAMMING/Intro_qCMB_class/2023_qCMB_class/Example_Data/winequality-red.csv", header = T)

new_winequality_df <- winequality_df %>% filter(quality > 6)
```

Export a dataframe to a .txt file, using write.table 
```{r}
write.table(new_winequality_df, file = '~/Library/CloudStorage/OneDrive-Colostate/JESSICA/CSU/BMB/ONISH_LAB/PROJECTS/PROGRAMMING/Intro_qCMB_class/2023_qCMB_class/Example_Data/new_winequality_df.txt', col.names = TRUE,
             row.names = FALSE, sep = "\t")
```

Export a dataframe to a .csv file, using write.table
```{r}
write.table(new_winequality_df, file = '~/Library/CloudStorage/OneDrive-Colostate/JESSICA/CSU/BMB/ONISH_LAB/PROJECTS/PROGRAMMING/Intro_qCMB_class/2023_qCMB_class/Example_Data/new_winequality_df.csv', col.names = TRUE,
             row.names = FALSE, sep = ",")
```

Now export the same dataframe to a .csv file, using write.csv 
```{r}
write.csv(new_winequality_df, file = '~/Library/CloudStorage/OneDrive-Colostate/JESSICA/CSU/BMB/ONISH_LAB/PROJECTS/PROGRAMMING/Intro_qCMB_class/2023_qCMB_class/Example_Data/new_winequality_df2.csv', row.names = FALSE)
```

An alternative to write.table (is meant to be faster)
```{r}
install.packages("data.table")
library(data.table)

fwrite(new_winequality_df, file = '~/Library/CloudStorage/OneDrive-Colostate/JESSICA/CSU/BMB/ONISH_LAB/PROJECTS/PROGRAMMING/Intro_qCMB_class/2023_qCMB_class/Example_Data/new_winequality_df2.txt', sep = "\t")

fwrite(new_winequality_df, file = '~/Library/CloudStorage/OneDrive-Colostate/JESSICA/CSU/BMB/ONISH_LAB/PROJECTS/PROGRAMMING/Intro_qCMB_class/2023_qCMB_class/Example_Data/new_winequality_df3.csv')
```

Another alternative to write.table
```{r}
library(readr)
write_tsv(new_winequality_df, path = '~/Library/CloudStorage/OneDrive-Colostate/JESSICA/CSU/BMB/ONISH_LAB/PROJECTS/PROGRAMMING/Intro_qCMB_class/2023_qCMB_class/Example_Data/new_winequality_df3.txt')

write_csv(new_winequality_df, path = '~/Library/CloudStorage/OneDrive-Colostate/JESSICA/CSU/BMB/ONISH_LAB/PROJECTS/PROGRAMMING/Intro_qCMB_class/2023_qCMB_class/Example_Data/new_winequality_df4.csv')
```

> Are all data file formats handled equally?

Lets check the reading speeds of the different functions 
```{r}
# Use system.time() to time how fast R processes an expression. 
# This function is especially useful for comparing the speed of two coding options.

system.time(
  read.csv("~/Library/CloudStorage/OneDrive-Colostate/JESSICA/CSU/BMB/ONISH_LAB/PROJECTS/PROGRAMMING/Intro_qCMB_class/2023_qCMB_class/Example_Data/winequality-red.csv", header = T)
)
```

```{r}
system.time(
  read.table("~/Library/CloudStorage/OneDrive-Colostate/JESSICA/CSU/BMB/ONISH_LAB/PROJECTS/PROGRAMMING/Intro_qCMB_class/2023_qCMB_class/Example_Data/winequality-red.csv", header = T, sep = ",")
)
```

```{r}
system.time(
  read.delim("~/Library/CloudStorage/OneDrive-Colostate/JESSICA/CSU/BMB/ONISH_LAB/PROJECTS/PROGRAMMING/Intro_qCMB_class/2023_qCMB_class/Example_Data/new_winequality_df.txt", header = T, sep = "\t")
)
```

What about writing speeds? 

* As a little home practice: 
    + Check to see which functions write data files faster 
    + Also see if particular file types write faster then others

> What if we have a really large data file?

* May run into R memory issues

* Check out the below resource for some extra reading into this topic 
    + [Out of memory strategies](https://bookdown.org/rdpeng/RProgDA/working-with-large-datasets.html)

