---
title: "Practice with Tidy Pipes and dataframes"
author: "David C. King"
date: "2023-02-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
library(tidyverse)
```

## Basic Piping with %>%

Pipes simplify the successive use of functions and provide an intuitive left-to-right order.

For example, here is the _mean_ of _differences_ between the first 8 numbers of the Fibonacci sequence.

```{r fib-mean-diff, eval=TRUE}
c(1,2,3,8,11) %>% diff() %>% mean()

# The order of operations is left-to-right, whereas the conventional form of the expression is nested:

mean(diff(c(1,2,3,8,11)))

```

### Breaking down by step

Before we had pipes, it was often necessary to save the output of each step in order to make the left-to-right order easier to see.

```{r, eval=TRUE}
# step 1
diffs = diff(c(1,2,3,8,11))
diffs

# step 2
mean_of_diffs = mean(diffs)
mean_of_diffs
```


### How it works

To understand what's happening, look at simple use cases.

`function(a)`

is the same as

`a %>% function()`

------------------------------------------------------------------------

If there is more than one argument to the function, the *additional* arguments are still listed in the function's parenthesis:

`function(a,b)`

is the same as

`a %>% function(b)`

and

`function(a,b,c)`

is the same as

`a %>% function(b,c)`

------------------------------------------------------------------------

### Practice

Practice with `sum()`

```{r practice-pipe-sum}
# write the pipe version of the following command:
sum(1,2) # 3

# pipe version: replace each '___' to make a valid statement. 
#___ %>% ___(___)


# To try your answer, delete the comment character in front of your pipe, and press the play button to execute this chunk. It should output '[1] 3' two times, once for each form of the statement.
```

Practice with `prod()`

```{r practice-pipe-prod}
# pipe version: replace each '___' to make a valid statement. 
5 %>% prod(10) # 50

# write the standard form of the previous command:
#___(___,___) 


#  To try your answer, delete the comment character in front of your new command, and press the play button to execute this chunk. It should output '[1] 50' two times, once for each form of the statement.
```

### Using vectors as values

A more common usage of pipes (and R in general), is to perform actions on collections of numbers in various types of objects like vectors or dataframes.

`vec = c(1,2,3)`

`vec %>% mean()`

Is the same as

`mean(vec)`

### Practice with vector variables

------------------------------------------------------------------------

```{r practice pipe with vector variables}
x = c(1,1,2,3,5,8)

# Take the mean of x: replace each '___' to make a valid statement. 
#___ %>% ___()

# write the non-pipe version of the previous command:
#___(___) 


# Take the standard deviation of x using the sd() function: replace each '___' to make a valid statement. 
# If you need help on using the sd() function, type ?sd in the console.
#___(___) 

# Write the pipe version of the previous command:
#___ %>% ___()


# Delete the comment character in front of your new commands and press the play button to execute this whole chunk. 
```

### Apply functions to variables in pipe form

1) Make some data and set it to x
2) Write the pipe form of several summary functions on x
3) Some summary functions: `mean`, `median`, `var`, `quantile`, `summary`

*Note: not all of these functions return a single value.

```{r from-scratch}
# Give yourself some data to work with by setting x to a vector (as above).
#x = ___

# write the tidyverse pipe form of the summary functions:

# x_mean = ...
# x_median = ...
# x_variance = ...
# x_quantile = ...
# x_summary = ...
```

### A note about assignment syntax

There are three valid operators to assign values to variables.  The following are equivalent.

1) `a = 3`

2) `a <- 3`

3) `3 -> a`

`a` is `3`

Different coders have different preferences, but the 3rd form in particular maintains the left-to-right order of operations in a pipe.

```{r variable-assignment}

x <- c(1,2,3,5,7,11,13,17) # some prime numbers
x
##### data %>% function() -> output

x %>% diff() -> diff_x
diff_x

##### data %>% step1() %>% step2() -> output

x %>% diff() %>% prod() -> product_of_prime_diffs
product_of_prime_diffs
```


## Piping output through multiple functions.

The __real__ benefit of piping is to apply functions one after another, as in the first example.

`c(1,2,3,8,11) %>% diff() %>% mean()`

`[1] 2.5`

More complex data lends itself to more useful examples- pipes are great for building, modifying, and analyzing data frames.

```{r}
# pretend RNA-seq data
df = data.frame(gene_names = c("geneA", "geneB", "geneC"),
           sampleA = c(0.3186023, 0.7902006, 0.4716855),
           sampleB = c(0.3218933, 0.5404638, 0.9218390))

df # print df to console
```
```{r}
# add a column to indicate the fold change
df %>% mutate(foldChange = sampleA/sampleB)
```

```{r}
# save the result to a new data frame
df2 = df %>% mutate(foldChange = sampleA/sampleB)
df2 # prints df2 to console
```

```{r}
# add a column called logFC that is the logarithm of foldChange
df3 = df2 %>% mutate(...)
df3 # prints df3 to console
```

```{r}
# Combine the two mutate steps above into a single pipe and store it as df4.
df4 = df %>% 
  mutate(...) %>%
  mutate(...) 

df4 # prints df4 to console        
```




### Prostate cancer study

<https://www.openintro.org/book/statdata/>

```{r prostate-data}
prostate = read.csv(url('https://www.openintro.org/book/statdata/exercise_prostate_cancer.csv'))
prostate
```

```{r summarize-prostate-columns}
# No grouping - overall mean of growth
# This is the same as taking the mean of the whole column "growth"
prostate_summary <- prostate %>% 
summarize(mean_growth=mean(growth))
prostate_summary

# serum only
# "after" serum seems to be associated with less growth, but this might obscure
# differences between cell types
prostate_summary <- prostate %>% 
  group_by(serum) %>%
  summarize(mean_growth=mean(growth))
prostate_summary

# cell type only
# These values are close to each other, and to the overall mean
prostate_summary <- prostate %>% 
  group_by(cell) %>%
  summarize(mean_growth=mean(growth))
prostate_summary

# cell type AND serum
# Effect on growth has different numeric ranges, and 
# may be different for different cell types
prostate_summary <- prostate %>% 
  group_by(serum,cell) %>%
  summarize(mean_growth=mean(growth))

prostate_summary %>% arrange(cell)
```

Add more summary fields

```{r add sd summary}
prostate_summary <- prostate %>%
  group_by(serum, cell) %>%
  summarize(mean_growth=...,sd_growth=...)
```

### If you're interested in the medical results
Is there a difference in group means between serum before/after, interactions are 
considered. Doing an ANOVA testing for different group means of cell type and serum,
yields an interesting result.

Differences between serum before/after are statistically robust, but 
immune vs. prostate cancer cells were not separate in magnitude
(probably not interesting to the authors)


## Practice with Kristin's DGE data


```{r data}
data <- read.delim("./DGE.tsv", sep="\t")[1:50,]
gene_lengths <- read.delim("./gene_ids.tsv", sep="\t")[1:50,]
```

### Merging

1) Show the non-tidy versions of merging data with gene_lengths.

2) Show results of different parameters- all.x versus all.y

3) Are there ways of comparing x not in y, y not in x.

```{r}


```


